---
title: "Pseudobulk sgRNA GEX effects CTR02"
author: "Eric Wang"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(grid)
library(pheatmap)
library(viridis)
library(scales)
library(DESeq2)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/scRNA_seq_analysis_functions.R")
source("functions/plotting_fxns.R")
theme_set(theme_Publication())
```

## [Data Import]{.underline}

```{r}
data <- readRDS("C:/Users/Eric/My Drive/Lab/datasets/EYW/CTR02_10x_240516/processing/CTR02_seurat_SCT_CRISPRumi9.rds")
```

Add some metadata columns

```{r}
data <- subset(data, subset = num_features == 1)
data$feature_gene <- gsub("\\..*","",data$feature_call)
# split hash ID into separate organ and CD62L status
hashSplit <- data@meta.data %>%
  dplyr::select(hash.ID) %>%
  separate(hash.ID,c("organ","CD62L_status"), sep = "-")
data$organ <- hashSplit$organ
data$CD62L_status <- hashSplit$CD62L_status
```

## [All Cells Pseudobulk]{.underline}

First, I will take a look at DEG analysis of pseudobulk analysis of all cells while adjusting for organ and CD62L status given their effects on transcriptional variability.

```{r, results='hide'}
# perform pseudobulk aggregation
pseudodata <- AggregateExpression(data, assays = "SCT", return.seurat = T, group.by = c("organ","CD62L_status","feature_call"))
pseudodata <- pseudodata[!grepl("^Tra[vdj]|^Trb[vdj]",rownames(pseudodata)),]

# gather info for metadata
# get UMIs per sgRNA
# get frequency of single sgRNA integrations for individual sgRNA
# get percent.mt
metaData <- data@meta.data %>%
  as_tibble(rownames = "cell_bc") %>%
  group_by(organ, CD62L_status, feature_gene, feature_call) %>%
  summarise(freq_cells = n()/nrow(data@meta.data),
            median_sgRNA_umi = median(as.numeric(num_umis)),
            mito_perc = mean(percent.mt))

# convert metadata to df form
ddsColData <- metaData  %>%
  mutate(feature_gene = gsub("NTC|NCC","control",feature_gene)) %>%
  as.data.frame()
rownames(ddsColData) <- paste(ddsColData$organ,ddsColData$CD62L_status,ddsColData$feature_call,sep="_")

# create DESeq2 object
dds <- DESeqDataSetFromMatrix(pseudodata@assays$SCT$counts,
                              colData = ddsColData,
                              design = ~ feature_gene + organ + CD62L_status)
dds <- DESeq(dds)

resLFCIfngr1 <- lfcShrink(dds, contrast = c("feature_gene","Ifngr1","control"), type = "ashr")
resLFCTgfbr2 <- lfcShrink(dds, contrast = c("feature_gene","Tgfbr2","control"), type = "ashr")

resLFCIfngr1Tidy <- resLFCIfngr1  %>%
  as_tibble(rownames = "genes") %>%
  arrange(padj)
resLFCTgfbr2Tidy <- resLFCTgfbr2  %>%
  as_tibble(rownames = "genes") %>%
  arrange(padj)

write_csv(resLFCIfngr1Tidy,"analysis_outs/pseudobulk_bulk_ifngr1_DEG.csv")
write_csv(resLFCTgfbr2Tidy,"analysis_outs/pseudobulk_bulk_tgfbr2_DEG.csv")
```

## [Ifngr1 KO comparison]{.underline}

Here, I will compare these results to the signature of Ifngr1 KO cells in https://www.nature.com/articles/s41590-023-01453-w, which used a flox mouse

```{r}
infectedDEG <- read_csv("external_datasets/gocher_infected_Ifngr1KO_Foxp3Cre_DEG.csv")
uninfectedDEG <- read_csv("external_datasets/gocher_uninfected_Ifngr1KO_Foxp3Cre_DEG.csv")
```

```{r, fig.height=12, fig.width=12}
infectedSigUp <- infectedDEG %>% filter(p_val_adj < 0.1 & avg_log2FC > 0)
infectedSigDown <- infectedDEG %>% filter(p_val_adj < 0.1 & avg_log2FC < 0)
uninfectedSigUp <- uninfectedDEG %>% filter(p_val_adj < 0.1 & avg_log2FC > 0)
uninfectedSigDown <- uninfectedDEG %>% filter(p_val_adj < 0.1 & avg_log2FC < 0)

ifngr1Comp <- resLFCIfngr1Tidy %>%
  mutate(infected_sig = case_when(genes %in% infectedSigUp$genes ~ "sig_up",
                                  genes %in% infectedSigDown$genes ~ "sig_down"),
         uninfected_sig = case_when(genes %in% uninfectedSigUp$genes ~ "sig_up",
                                  genes %in% uninfectedSigDown$genes ~ "sig_down")) %>%
  mutate(infected_sig = replace_na(infected_sig,"not_sig"),
         uninfected_sig = replace_na(uninfected_sig, "not_sig")) %>%
  mutate(infected_sig = factor(infected_sig, c("sig_up","sig_down","not_sig")),
         uninfected_sig = factor(uninfected_sig, c("sig_up","sig_down","not_sig"))) %>%
  filter(padj < 0.1)

Tgfbr2Comp <- resLFCTgfbr2Tidy %>%
  mutate(infected_sig = case_when(genes %in% infectedSigUp$genes ~ "sig_up",
                                  genes %in% infectedSigDown$genes ~ "sig_down"),
         uninfected_sig = case_when(genes %in% uninfectedSigUp$genes ~ "sig_up",
                                  genes %in% uninfectedSigDown$genes ~ "sig_down")) %>%
  mutate(infected_sig = replace_na(infected_sig,"not_sig"),
         uninfected_sig = replace_na(uninfected_sig, "not_sig")) %>%
  mutate(infected_sig = factor(infected_sig, c("sig_up","sig_down","not_sig")),
         uninfected_sig = factor(uninfected_sig, c("sig_up","sig_down","not_sig"))) %>%
  filter(padj < 0.1)

p1 <- ggplot(ifngr1Comp , aes(x = log2FoldChange, y = -log10(padj), color = infected_sig)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  xlim(-2.3,2.3) +
  theme(aspect.ratio = 1) +
  ggtitle("Perturb Ifngr1 KO DEGs\n(pseudobulk)")

p2 <- ggplot(ifngr1Comp , aes(x = log2FoldChange, y = -log10(padj), color = uninfected_sig)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  xlim(-2.3,2.3) +
  theme(aspect.ratio = 1) +
  ggtitle("Perturb Ifngr1 KO DEGs\n(pseudobulk)")

p3 <- ggplot(Tgfbr2Comp , aes(x = log2FoldChange, y = -log10(padj), color = infected_sig)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  xlim(-2.3,2.3) +
  theme(aspect.ratio = 1) +
  ggtitle("Perturb Tgfbr2 KO DEGs\n(pseudobulk)")

p4 <- ggplot(Tgfbr2Comp , aes(x = log2FoldChange, y = -log10(padj), color = uninfected_sig)) +
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  xlim(-2.3,2.3) +
  theme(aspect.ratio = 1) +
  ggtitle("Perturb Tgfbr2 KO DEGs\n(pseudobulk)")

(p1+p2)/(p3+p4)
ggsave("analysis_outs/perturb_ifngr1flox_comp_volcano.pdf")
```

```{r, fig.height=10, fig.width=12}
p1 <- resLFCIfngr1Tidy %>%
  left_join(infectedDEG) %>%
  filter(padj < 0.1 | p_val_adj < 0.1) %>%
  drop_na() %>%
  mutate(sig_category = case_when(padj < 0.1 & p_val_adj < 0.1 ~ "both",
                                  padj >= 0.1 & p_val_adj < 0.1 ~ "infected_only",
                                  padj < 0.1 & p_val_adj >= 0.1 ~ "perturb_only",)) %>%
  mutate(sig_category = factor(sig_category, c("both","perturb_only","infected_only"))) %>%
  ggplot(aes(x = log2FoldChange, y = avg_log2FC, color = sig_category)) +
    geom_point()+
    scale_color_brewer(palette = "Dark2") +
    xlim(-4,4) +
    ylim(-4,4) +
    theme(aspect.ratio = 1) +
    ggtitle("Ifngr1 KO comparison")

p2 <- resLFCIfngr1Tidy %>%
  left_join(uninfectedDEG) %>%
  filter(padj < 0.1 | p_val_adj < 0.1) %>%
  drop_na() %>%
  mutate(sig_category = case_when(padj < 0.1 & p_val_adj < 0.1 ~ "both",
                                  padj >= 0.1 & p_val_adj < 0.1 ~ "uninfected_only",
                                  padj < 0.1 & p_val_adj >= 0.1 ~ "perturb_only",)) %>%
  mutate(sig_category = factor(sig_category, c("both","perturb_only","uninfected_only"))) %>%
  ggplot(aes(x = log2FoldChange, y = avg_log2FC, color = sig_category)) +
    geom_point()+
    scale_color_brewer(palette = "Dark2") +
    xlim(-4,4) +
    ylim(-4,4) +
    theme(aspect.ratio = 1) +
    ggtitle("Ifngr1 KO comparison")

p3 <- resLFCTgfbr2Tidy %>%
  left_join(infectedDEG) %>%
  filter(padj < 0.1 | p_val_adj < 0.1) %>%
  drop_na() %>%
  mutate(sig_category = case_when(padj < 0.1 & p_val_adj < 0.1 ~ "both",
                                  padj >= 0.1 & p_val_adj < 0.1 ~ "infected_only",
                                  padj < 0.1 & p_val_adj >= 0.1 ~ "perturb_only",)) %>%
  mutate(sig_category = factor(sig_category, c("both","perturb_only","infected_only"))) %>%
  ggplot(aes(x = log2FoldChange, y = avg_log2FC, color = sig_category)) +
    geom_point()+
    scale_color_brewer(palette = "Dark2") +
    xlim(-4,4) +
    ylim(-4,4) +
    theme(aspect.ratio = 1) +
    ggtitle("Tgfbr2 KO comparison")

p4 <- resLFCTgfbr2Tidy %>%
  left_join(infectedDEG) %>%
  filter(padj < 0.1 | p_val_adj < 0.1) %>%
  drop_na() %>%
  mutate(sig_category = case_when(padj < 0.1 & p_val_adj < 0.1 ~ "both",
                                  padj >= 0.1 & p_val_adj < 0.1 ~ "uninfected_only",
                                  padj < 0.1 & p_val_adj >= 0.1 ~ "perturb_only",)) %>%
  mutate(sig_category = factor(sig_category, c("both","perturb_only","uninfected_only"))) %>%
  ggplot(aes(x = log2FoldChange, y = avg_log2FC, color = sig_category)) +
    geom_point()+
    scale_color_brewer(palette = "Dark2") +
    xlim(-4,4) +
    ylim(-4,4) +
    theme(aspect.ratio = 1) +
    ggtitle("Tgfbr2 KO comparison")

(p1+p2)/(p3+p4)
```


## [Tissue and activation status adjustment]{.underline}

Here, I will compare the DEG analysis results with and without accounting for cofactors in DESeq2 (e.g. tissue of origin and activation status)

```{r, results="hide"}
pseudobulk <- function(x,covariates){
  # perform pseudobulk aggregation
  pseudodata1 <- AggregateExpression(x, assays = "SCT", return.seurat = T, group.by = covariates)
  pseudodata1 <- pseudodata1[!grepl("^Tra[vdj]|^Trb[vdj]",rownames(pseudodata1)),]
  pseudodata1$feature_gene <- gsub("NTC|NCC","control",pseudodata1$feature_gene) %>%
    factor(c("control","Ifngr1","Tgfbr2"))
  
  return(pseudodata1)
}

pseudodata1 <- pseudobulk(data, c("organ","CD62L_status","feature_gene","feature_call"))
pseudodata2 <- pseudobulk(data, c("organ","feature_gene","feature_call"))
pseudodata3 <- pseudobulk(data, c("CD62L_status","feature_gene","feature_call"))
pseudodata4 <- pseudobulk(data, c("feature_gene","feature_call"))

dds <- vector(mode = "list")
dds[["organ_CD62L"]] <- DESeqDataSetFromMatrix(pseudodata1@assays$SCT$counts,
                                                               colData = pseudodata1@meta.data,
                                                               design = ~ organ + CD62L_status + feature_gene)
dds[["organ"]] <- DESeqDataSetFromMatrix(pseudodata2@assays$SCT$counts,
                                                               colData = pseudodata2@meta.data,
                                                               design = ~ organ + feature_gene)
dds[["CD62L"]] <- DESeqDataSetFromMatrix(pseudodata3@assays$SCT$counts,
                                                               colData = pseudodata3@meta.data,
                                                               design = ~ CD62L_status + feature_gene)
dds[["none"]] <- DESeqDataSetFromMatrix(pseudodata4@assays$SCT$counts,
                                                               colData = pseudodata4@meta.data,
                                                               design = ~ feature_gene)

# run DEseq analysis
dds <- lapply(dds, function(x) DESeq(x))
```

First, we'll look at Ifngr1 KO DEGs

```{r, fig.height=5, fig.width=12}
# calculate DEGs for all conditions
ifngr1DEG <- function(x){
  resLFC <- lfcShrink(x, contrast = c("feature_gene","Ifngr1","control"), type = "ashr")
  resLFC <- resLFC %>%
    as_tibble(rownames = "genes")
}

# create list of tibbles with DEG results
resIfngr1 <- lapply(dds, function(x) ifngr1DEG(x))

# Combine the tibbles into a single data frame with an identifier
combinedTib <- bind_rows(
  lapply(seq_along(resIfngr1), function(i) {
    resIfngr1[[i]] %>%
      mutate(tibble_id = names(resIfngr1)[i])
  })
)

# Separate the first tibble for comparison
referenceTib <- combinedTib %>% filter(tibble_id == "none")
comparisonTib <- combinedTib %>% filter(tibble_id != "none")

# Merge the reference tibble with the comparison data
comparisonTib1 <- comparisonTib %>%
  left_join(referenceTib, by = "genes", suffix = c("", "_ref")) %>%
  mutate(
    significance = case_when(
      padj_ref < 0.1 & padj < 0.1 ~ "both",
      padj_ref < 0.1 ~ "no adjustment specific",
      padj < 0.1 ~ "adjustment specific",
      TRUE ~ "n.s."
    )) %>%
  mutate(significance = factor(significance, c("n.s.","no adjustment specific","adjustment specific","both"))
  )

# Plotting
ggplot(comparisonTib1, aes(x = log2FoldChange_ref, y = log2FoldChange, color = significance)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~tibble_id) +
  theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  xlim(-3,3) +
  ylim(-3,3) +
  labs(
    title = "Comparison of shrunken Log2FC for Ifngr1 KO",
    x = "log2FC no covariates",
    y = "log2FC covariates"
  )
```

```{r, fig.height=5, fig.width=12}
# calculate DEGs for all conditions
tgfbr2DEG <- function(x){
  resLFC <- lfcShrink(x, contrast = c("feature_gene","Tgfbr2","control"), type = "ashr")
  resLFC <- resLFC %>%
    as_tibble(rownames = "genes")
}

# create list of tibbles with DEG results
resTgfbr2 <- lapply(dds, function(x) tgfbr2DEG(x))

# Combine the tibbles into a single data frame with an identifier
combinedTib <- bind_rows(
  lapply(seq_along(resTgfbr2), function(i) {
    resTgfbr2[[i]] %>%
      mutate(tibble_id = names(resTgfbr2)[i])
  })
)

# Separate the first tibble for comparison
referenceTib <- combinedTib %>% filter(tibble_id == "none")
comparisonTib <- combinedTib %>% filter(tibble_id != "none")

# Merge the reference tibble with the comparison data
comparisonTib2 <- comparisonTib %>%
  left_join(referenceTib, by = "genes", suffix = c("", "_ref")) %>%
  mutate(
    significance = case_when(
      padj_ref < 0.1 & padj < 0.1 ~ "both",
      padj_ref < 0.1 ~ "no adjustment specific",
      padj < 0.1 ~ "adjustment specific",
      TRUE ~ "n.s."
    )) %>%
  mutate(significance = factor(significance, c("n.s.","no adjustment specific","adjustment specific","both"))
  )

# Plotting
ggplot(comparisonTib2, aes(x = log2FoldChange_ref, y = log2FoldChange, color = significance)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~tibble_id) +
  theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  xlim(-3,3) +
  ylim(-3,3) +
  labs(
    title = "Comparison of shrunken Log2FC for Tgfbr2 KO",
    x = "log2FC no covariates",
    y = "log2FC covariates"
  )
```

## [TCR clonal removal]{.underline}

Here, I will compare the DEG analysis results with and without accounting for cofactors in DESeq2 (e.g. tissue of origin and activation status)

```{r, results="hide"}
pseudobulk <- function(x,covariates){
  # perform pseudobulk aggregation
  pseudodata1 <- AggregateExpression(x, assays = "SCT", return.seurat = T, group.by = covariates)
  pseudodata1 <- pseudodata1[!grepl("^Tra[vdj]|^Trb[vdj]",rownames(pseudodata1)),]
  pseudodata1$feature_gene <- gsub("NTC|NCC","control",pseudodata1$feature_gene) %>%
    factor(c("control","Ifngr1","Tgfbr2"))
  
  return(pseudodata1)
}

data2 <- subset(data, subset = clonalFrequency < 5)
data3 <- subset(data, subset = clonalFrequency <= 1)

pseudodata1 <- pseudobulk(data, c("organ","CD62L_status","feature_gene","feature_call"))
pseudodata2 <- pseudobulk(data2, c("organ","CD62L_status","feature_gene","feature_call"))
pseudodata3 <- pseudobulk(data2, c("organ","CD62L_status","feature_gene","feature_call"))

dds <- vector(mode = "list")
dds[["full"]] <- DESeqDataSetFromMatrix(pseudodata1@assays$SCT$counts,
                                                               colData = pseudodata1@meta.data,
                                                               design = ~ organ + CD62L_status + feature_gene)
dds[["clone_remove_5"]] <- DESeqDataSetFromMatrix(pseudodata2@assays$SCT$counts,
                                                               colData = pseudodata2@meta.data,
                                                               design = ~ organ + CD62L_status + feature_gene)
dds[["clone_remove_1"]] <- DESeqDataSetFromMatrix(pseudodata3@assays$SCT$counts,
                                                               colData = pseudodata3@meta.data,
                                                               design = ~ organ + CD62L_status + feature_gene)

# run DEseq analysis
dds <- lapply(dds, function(x) DESeq(x))
```

First, we'll look at Ifngr1 KO DEGs

```{r, fig.height=5, fig.width=12}
# calculate DEGs for all conditions
ifngr1DEG <- function(x){
  resLFC <- lfcShrink(x, contrast = c("feature_gene","Ifngr1","control"), type = "ashr")
  resLFC <- resLFC %>%
    as_tibble(rownames = "genes")
}

# create list of tibbles with DEG results
resIfngr1 <- lapply(dds, function(x) ifngr1DEG(x))

# Combine the tibbles into a single data frame with an identifier
combinedTib <- bind_rows(
  lapply(seq_along(resIfngr1), function(i) {
    resIfngr1[[i]] %>%
      mutate(tibble_id = names(resIfngr1)[i])
  })
)

# Separate the first tibble for comparison
referenceTib <- combinedTib %>% filter(tibble_id == "full")
comparisonTib <- combinedTib %>% filter(tibble_id != "full")

# Merge the reference tibble with the comparison data
comparisonTib1 <- comparisonTib %>%
  left_join(referenceTib, by = "genes", suffix = c("", "_ref")) %>%
  mutate(
    significance = case_when(
      padj_ref < 0.1 & padj < 0.1 ~ "both",
      padj_ref < 0.1 ~ "no adjustment specific",
      padj < 0.1 ~ "adjustment specific",
      TRUE ~ "n.s."
    )) %>%
  mutate(significance = factor(significance, c("n.s.","no adjustment specific","adjustment specific","both"))
  )

# Plotting
ggplot(comparisonTib1, aes(x = log2FoldChange_ref, y = log2FoldChange, color = significance)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~tibble_id) +
  theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  xlim(-3,3) +
  ylim(-3,3) +
  labs(
    title = "Comparison of shrunken Log2FC for Ifngr1 KO",
    x = "log2FC full",
    y = "log2FC adjustment"
  )
```

```{r, fig.height=5, fig.width=12}
# calculate DEGs for all conditions
tgfbr2DEG <- function(x){
  resLFC <- lfcShrink(x, contrast = c("feature_gene","Tgfbr2","control"), type = "ashr")
  resLFC <- resLFC %>%
    as_tibble(rownames = "genes")
}

# create list of tibbles with DEG results
resTgfbr2 <- lapply(dds, function(x) tgfbr2DEG(x))

# Combine the tibbles into a single data frame with an identifier
combinedTib <- bind_rows(
  lapply(seq_along(resTgfbr2), function(i) {
    resTgfbr2[[i]] %>%
      mutate(tibble_id = names(resTgfbr2)[i])
  })
)

# Separate the first tibble for comparison
referenceTib <- combinedTib %>% filter(tibble_id == "full")
comparisonTib <- combinedTib %>% filter(tibble_id != "full")

# Merge the reference tibble with the comparison data
comparisonTib2 <- comparisonTib %>%
  left_join(referenceTib, by = "genes", suffix = c("", "_ref")) %>%
  mutate(
    significance = case_when(
      padj_ref < 0.1 & padj < 0.1 ~ "both",
      padj_ref < 0.1 ~ "no adjustment specific",
      padj < 0.1 ~ "adjustment specific",
      TRUE ~ "n.s."
    )) %>%
  mutate(significance = factor(significance, c("n.s.","no adjustment specific","adjustment specific","both"))
  )

# Plotting
ggplot(comparisonTib2, aes(x = log2FoldChange_ref, y = log2FoldChange, color = significance)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~tibble_id) +
  theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  xlim(-3,3) +
  ylim(-3,3) +
  labs(
    title = "Comparison of shrunken Log2FC for Tgfbr2 KO",
    x = "log2FC full",
    y = "log2FC adjustment")
```

